---
title: "Transmission effect on fuel consumption"
author: "Balint SZEBENYI"
date: "2015. augusztus 13."
output: html_document
---

```{r, echo = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(reshape)
data(mtcars)
```

#Executive summary

```{r}
summary(cars)
```

# Exploratory analysis
The dataset, which was examined is the mtcars dataset that is distributed with R. The original data description is found at <https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html>, this is only a brief summary.

There are eleven variables in the dataset, the two relevant ones are:
* mpg - Fuel consumption, miles in gallon
* am - Transmission: automatic (1) or manual (0) 

'* cyl - Number of cylinders
'* disp - Displacement in qubic inches
'* hp - Gross horsepower
'* wt - Weight in 1000 lbs

The relationship between the two variables can be illustrated via the following plot.

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
